Kindly Bent to Free Us

### 2019-11-21 kindly/motivation.tex
Iter Example: Removed empty newlines before lstlistings
Rewrite the intro example.

\lang{} is amenable to a functional resource-passing style
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement). But it really shines 
when manipulating mutable resources like buffers or connection pools
using a mixed functional and imperative programming style.
%
To support this usage pattern of linearity, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
we demonstrate through a series of examples:
\begin{itemize}
\item An API for linear arrays to demonstrate simple borrowing and imperative programming.
\item
  A backtracking Sudoku solver to demonstrate
  \lang's ability to build safe hybrid functional/imperative data-structures.
\item A pool of linear objects to demonstrate advanced use of regions.
\end{itemize}

\subsection{Imperative programming with linear arrays}
\label{sec:imper-progr}

The API for mutable linear arrays (shown in \cref{sig:array})
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use @create n v@ to create
an (unrestricted) array of size @n@ initialized with value
@v@. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using @map@ (or @set@ if
affine). The @get@ function is only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like @create@ and @free@ manipulate the whole
array. %As with channels in \cref{sec:session-linearity},
The constructor
@create@ yields a linear resource which is consumed by @free@.
During the lifetime of the array resource @a@, we can split
off \emph{shared borrows} @&a@ that provide a read-only
view or \emph{exclusive borrows} @&!a@ for read-write views.
Observer functions such as @length@ and @get@ expect an shared borrow
argument, mutator functions such a @set@ expect
a exclusive borrow.

@@

The code in \cref{ex:array} demonstrates a use of the API to create an
array of Fibonacci numbers as the programmer would write it. After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} of the example contains two shared borrows
in the same expression, which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set}, which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
module type ARRAY = sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int -> 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set : ('a : aff) => &!('a t) * int * 'a -> unit
  val map : (&'a -> 'b) -> &('a t) -> 'b t
end

module Array : ARRAY
\end{lstlisting}
    \vspace{-15pt}
    \caption{Signature}
    \label{sig:array}
  \end{subfigure}

  \begin{subfigure}[t]{1\linewidth}
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}
    \vspace{-10pt}
    \caption{Example of use}
    \label{ex:array}
\begin{lstlisting}[firstnumber=2]
  let a = create n 1 in
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}
    \vspace{-10pt}
    \caption{Excerpt of \cref{ex:array} with explicit regions}
    \label{ex:array:region}
  \end{subfigure}
  \vspace{-5pt}
  \caption{Linear arrays}
  \label{ex:array}
\end{figure}

\cref{ex:array:region} uses braces @{| ... |}@ to make the
regions of the example in \cref{ex:array} explicit.  One
region consists of the header expression of the @let@ in
\cref{ex:array:region:get}. It is contained in another region that
comprises the body of the @for@ loop. \lang guarantees that borrows
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like @aff_r@ and @un_r@ where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of @&!a@ in
\cref{ex:array:set} has kind @aff_1@ whereas the type of
@&a@ in \cref{ex:array:region:get} has kind
@un_2@ and the typing of the inner region is such that types with
kind indexes greater than or equal to $2$ cannot escape.
In the example, borrows cannot escape  because they are consumed
immediately by @get@ and @set@.

Region annotations are automatically inserted using
syntactic cues such as borrows and binders (see
\cref{regionannot}).  Users may manually insert regions to further restrict the range of a
borrow. In practice, region annotations are
rarely needed with common programming idioms.
% All regions for the examples in this section have been inferred.

% \begin{figure}
%   \centering
%   % \begin{subfigure}{0.6\linewidth}
% \begin{lstlisting}
% let mk_fib_array n =
%   let a = create n 1 in
%   for i = 2 to n - 1 do {|
%     let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
%     set &!a i x
%   |}
%   done;
%   a(*@\label{ex:array:return}*)
% # mk_fib_array : int -> int Array.t
% \end{lstlisting}
%     \caption{Example from \cref{ex:array} with explicit regions}
%     \label{ex:array:region}
%   % \end{subfigure}
% \end{figure}

@@

Recently introduced persistent data structures
permit transient mutations where
non-linear uses lead to degraded performance
\cite{DBLP:conf/ml/ConchonF07} or to
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
APIs in several non-pure functional languages (OCaml, Clojure, \dots).
Affine types help formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an API for immutable data types.
%

In this section, we present
a safe API for persistent arrays that support both immutable and mutable features,
and use it to implement a backtracking Sudoku solver.
% show how \lang allows us to define
The Sudoku solver maintains an array to represent the state of
the game and uses backtracking when there are several choices to proceed.
The use of backtracking suggests a persistent data structure for the array.
However, only changes that correspond to a choice point need to use
the persistence mechanism, others may be implemented as
cheap in-place mutations.

\cref{sig:hybarray} contains an API @HybArray@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the @Array@
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of @Array@.
The in-place mutation function @set_mut@ with type
@&!('a t) * int * 'a -> unit@ works on a exclusive borrow  whereas the persistent
@set@ operation has type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
@HybArray@ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
@mapi: (int -> &'a -> 'b) -> &('a t) -> 'b t@
is a simple variation on @Array.map@ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

\begin{figure}[tp]
  \centering
  % \begin{subfigure}{1\linewidth}
\begin{lstlisting}
module type HybArray = sig
  include ARRAY
  val set : &('a t) * int * 'a -> 'a t
  val set_mut : &!('a t) * int * 'a -> unit
end

module Cow : HybArray = struct
  include Array
  let set (a, i0, x0) =
    Array.mapi (fun i x -> if i = i0 then x0 else x) a
  let set_mut = Array.set
end
\end{lstlisting}
    \vspace{-15pt}
  %   \caption{Implementation using copy-on-write}
  % \end{subfigure}
  \caption{Hybrid arrays}
    \label{sig:hybarray}
    \label{ex:cow}

  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
    \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}

@@

While presented for copy-on-write arrays, the API
can easily be adapted to other persistent data structures with
transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
convenient programming style without compromising performance.

@@

However, if we want to work on linear and affine resources,
we must make sure those resources are not leaked in the result.
content. We obtain the following signature:

\begin{lstlisting}
val fold :
  ('b:'k),('k <= aff_r) =>
  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
\end{lstlisting}

@@

We can then easily use this function to fold over an array of files. Here
we compute the sum of their sizes:

\begin{lstlisting}
let total_size = fold (fun f s -> File.size f + s) &all_files 0 
\end{lstlisting}

Naturally, this is not sufficient if we want to mutate the array during iteration.
For this purpose, we need to take an exclusive borrow of the structure to iterate
on:

\begin{lstlisting}
val iter_mut : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

Since the callback will be called multiple times, it is explicitly
annotated with @un@.

While the distinction between mutable and immutable iteration function
might seem unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It allows the programmer to explicitly state how various iterations
are allowed to compose and be optimized.
\lang's region variables ensures that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use-cases of linearity such
as pools of linear objects (see \cref{tuto:pool}).

>>>>

\lang{} is amenable to a functional resource-passing style
common in functional encodings of session types (e.g.,
\cite{DBLP:journals/jfp/Padovani17}; see also
\cref{sec:session-linearity} in the supplement). But it really shines
when manipulating mutable resources like buffers or connection pools
using a mixed functional and imperative programming style.
%
To support this usage pattern of linearity, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05} which
we demonstrate through a series of examples.

As running theme of our examples, we use an API for persistent arrays.
Recently introduced persistent data structures
permit transient mutations where
non-linear uses lead to degraded performance
\cite{DBLP:conf/ml/ConchonF07} or to
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Tries (HAMT) have been used with similar
APIs in several non-pure functional languages (OCaml, Clojure, \dots).
Affine types help formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an API for immutable data types.

\cref{sig:hybarray} contains an API @PersistArray@
that enables using mutable and immutable modifications to
the board through affine types and borrows.
%
We first present how to use it imperatively, as if it was a simple
mutable linear array. We then see how to exploit its persistent
properties to write a backtracking sudoku solver. Finally, we consider
iterators on such linear structures.

% \begin{itemize}
% \item An API for linear arrays to demonstrate simple borrowing and imperative programming.
% \item
%   A backtracking Sudoku solver to demonstrate
%   \lang's ability to build safe hybrid functional/imperative data-structures.
% \item A pool of linear objects to demonstrate advanced use of regions.
% \end{itemize}

\begin{figure}[tp]
  \centering
  % \begin{subfigure}{1\linewidth}
\begin{lstlisting}
module PersistArray : sig
  type ('a : 'k) t : lin
  val create : ('a : un) => int -> 'a -> 'a t
  val free : ('a : aff) => 'a t -> unit
  val length : &('a t) -> int
  val get : ('a : un) => &('a t) * int -> 'a
  val set_mut : ('a : aff) => &!('a t) * int * 'a -> unit
  val set : ('a : un) => &('a t) * int * 'a -> 'a t
  val map : (&'a -> 'b) -> &('a t) -> 'b t
end
\end{lstlisting}
% \begin{lstlisting}
% (* A copy-on-write implementation *)
% module Cow : PersistArray = struct
%   type 'a t = 'a array
%   ...
%   let set (a, i0, x0) =
%     mapi (fun i x -> if i = i0 then x0 else x) a
% end
% \end{lstlisting}
  \vspace{-15pt}
  % \caption{Implementation using copy-on-write}
  % \end{subfigure}
  \caption{Hybrid arrays}
  \label{sig:array}
  \label{sig:hybarray}
  \label{ex:cow}
\end{figure}

\subsection{Imperative programming}

Our API for linear arrays
aims to safely handle manual allocation and
deallocation of arrays that may contain affine elements.
One would first use @create n v@ to create
an (unrestricted) array of size @n@ initialized with value
@v@. This value must be unrestricted as it is duplicated to
initialize all array elements. Using the @map@ function we
can transform the unrestricted elements into linear (affine)
ones. This array can be processed further using @map@ (or @set_mut@ if
affine). The @get@ and @set@ function are only applicable if the element type
is unrestricted as one element is duplicated. The @length@
function is always applicable. To @free@ an array the elements must be
affine.

To manage the different kinds of accessing the array we distinguish between constructors, destructors,
observers, and mutators.
Constructors and destructors like @create@ and @free@ manipulate the whole
array. %As with channels in \cref{sec:session-linearity},
The constructor
@create@ yields a linear resource which is consumed by @free@.
During the lifetime of the array resource @a@, we can split
off \emph{shared borrows} @&a@ that provide a read-only
view or \emph{exclusive borrows} @&!a@ for read-write views.
Observer functions such as @length@, @get@ and @set@ expect a shared borrow
argument, mutator functions such a @set_mut@ expect
a exclusive borrow.

@@

We can use these arrays in a purely mutable fashion
to create an
array of Fibonacci numbers using code very similar to an usual imperative style:
\begin{lstlisting}
let mk_fib_array n =
  let a = create n 1 in
  for i = 2 to n - 1 do
    let x = get (&a, i-1) + get (&a, i-2) in(*@\label{ex:array:get}*)
    set_mut (&!a, i, x)(*@\label{ex:array:set}*)
  done;
  a(*@\label{ex:array:return}*)
# mk_fib_array : int -> int Array.t
\end{lstlisting}

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} of the example contains two shared borrows
in the same expression, which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set}, which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.

More precisely, the code below is an annotated excerpt with explicit regions
marked in braces @{| ... |}@:

\begin{lstlisting}[firstnumber=3]
  for i = 2 to n - 1 do {|
    let x = {| get &a (i-1) + get &a (i-2) |} in(*@\label{ex:array:region:get}*)
    set &!a i x
  |} done;
\end{lstlisting}

One
region consists of the header expression of the @let@ in
\cref{ex:array:region:get}. It is contained in another region that
comprises the body of the @for@ loop. \lang guarantees that borrows
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like @aff_r@ and @un_r@ where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of @&!a@ in
\cref{ex:array:set} has kind @aff_1@ whereas the type of
@&a@ in \cref{ex:array:region:get} has kind
@un_2@ and the typing of the inner region is such that types with
kind indexes greater than or equal to $2$ cannot escape.
In the example, borrows cannot escape  because they are consumed
immediately by @get@ and @set_mut@.

@@

\begin{figure}[tp]
  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
  \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}

On top of the mutation function @set_mut@, our array implementation
also provides @set@ of type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh, modified structure.
%
This API could be implemented in a wide variety of ways, such as
copy-on-write (which copies the whole array on each @set_mut@)
or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.

We can use our persistent arrays
to implement a backtracking Sudoku solver.
% show how \lang allows us to define
The Sudoku solver maintains an array to represent the state of
the game and uses backtracking when there are several choices to proceed.
The use of backtracking suggests a persistent data structure for the array.
However, only changes that correspond to a choice point need to use
the persistence mechanism, others may be implemented as
cheap in-place mutations.

@@

While presented for persistent arrays, this API
can easily be adapted to other persistent data structures with
transient mutability such as Relaxed-Radix Balance Vectors (RRB) \cite{DBLP:journals/pacmpl/Puente17}
or persistent HAMTs \cite{bagwell2001ideal,clojurehamt} to provide  a
convenient programming style without compromising performance.

@@

However, if we want to work on linear and affine resources,
we must make sure those resources are not leaked in the result.
We obtain the following signature:
\begin{lstlisting}
val fold :
  ('b:'k),('k <= aff_r) =>
  (&(aff_r1,'a) -> 'b -{aff_r1}> 'b) -> &('k_1,'a array) -> 'b -{'k_1}> 'b
\end{lstlisting}

@@

We can then easily use this function to fold over an array of files. Here
we compute the sum of their sizes:
\begin{lstlisting}
let total_size = fold (fun f s -> File.size f + s) &all_files 0
\end{lstlisting}

Naturally, this is not sufficient if we want to mutate the array during iteration,
as is often the case with @iter@ functions.
For this purpose, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}
val iter : &!('a array) -> (&!(aff_r1,'a) -{un}> unit) -{aff}> unit
\end{lstlisting}

While the distinction between mutable and immutable iteration functions
might seem unfortunate, it is typical of
programming with borrows and is also present in the Rust standard library.
It allows the programmer to explicitly state how various iterations
are allowed to compose and be optimized.
\lang's region variables ensures that the content iterated on can never
be leaked outside of the iteration function.
This pattern is essential in many use-cases of linearity such
as pools of linear objects (see example in \cref{tuto:pool}).
